Model Finding for Recursive Functions in SMT

### 2016-03-30 ijcar2016/body.tex
TOdid some more

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying our translation, described in
Section~\ref{sec:encoding}. It is useful to connect it to the
standard notion of \emph{well-founded} function definitions, often called
\emph{terminating} definitions in a slight abuse of terminology. In such
definitions, all recursive calls are decreasing with respect to a well-founded
relation, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well founded, it is sufficient to require that
the defined function be terminating when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\,
\con{p}( x ) \teq \lite\bigl( x \leq 0,\allowbreak\;  1,\allowbreak\;
  2 * \con{p}( x - 1 ) \bigr)$,
where the theory $T$ is integer arithmetic extended with the %uninterpreted
symbol $\con p:\typeint \rightarrow \typeint$, can be shown to be well founded
under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

@@

Another useful class of function definitions is that of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be ill founded, without
their being inconsistent. 
Intuitively, productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the function defined by
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
%%% If what follows ':' is a complete sentence, the standard Am.E. convention
%%% is to capitalize it. --JB
falls within this class: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specifies the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that unfolding
a corecursive definition infinitely computes a unique infinite object.
%We conjecture that a large fragment of mixed recursive--corecursive
%specifications are

\begin{rep}
Beyond totality, an admissible set can contain underspecified functions
such as $\forallf{\con{f}} x : \typeint.\allowbreak\;\, \con{f}( x )
\teq \con{f}( x - 1 )$ or $\forallf{\con{g}} x.\allowbreak\;\, \con{g}( x
) \teq \con{g}( x )$. We conjecture that one can ignore all
tail-recursive calls (i.e., calls that occupy the right-hand side of the
definition, potentially under some $\lite$ branch) when establishing well-foundedness
or productivity, without affecting admissibility.
\end{rep}

@@

An example of an inadmissible set is
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$,
where $T$ is integer arithmetic extended to a set of uninterpreted symbols
$\{\con f, \con g :\typeint \rightarrow \typeint,\, \ldots \}$.
The set is inadmissible because the formula $\ltrue$ is
closed under function expansion with respect to this set
(since $\con{f}$ does not occur in $\ltrue$),
and yet there is no model of $T$
satisfying $\con{f}$'s definition. A more subtle example is
$\badassex$.
While this set has a model where $\con f$ and $\con g$ are interpreted as the
constant function $0$, it is not admissible since %the formula
$\con{f}( 0 ) \teq 1$ is
closed under function expansion \begin{rep}with respect to this set\end{rep}
and yet there exists
no interpretation satisfying both $\con{f}( 0 ) \teq 1$ and $\con{g}$'s
definition.

@@

The input syntax was inspired by that other systems based on higher-order
logic (e.g., Isabelle/HOL) and by functional programming languages (e.g.,
OCaml). The following simple problem gives a taste of the syntax:
%
\begin{quote}
\verb$data nat := 0 | Suc nat.$ \\[2\jot]
\verb$pred even : nat -> prop :=$ \\
\verb$  even 0;$ \\
\verb$  forall n. odd n => even (Suc n)$ \\
\verb$and odd : nat -> prop :=$ \\
\verb$  forall n. even n => odd (Suc n).$ \\[2\jot]
\verb$val m : nat.$ \\
\verb$goal even m && ~ (m = 0).$
\end{quote}
%
The problem defines a datatype (\texttt{nat}) and two mutually recursive
inductive predicates (\texttt{even} and \texttt{odd}), it declares a constant
\texttt{m}, and specifies a goal to satisfy (``\texttt{m} is even and nonzero'').
Nunchaku quickly finds the following partial model:
%
\begin{quote}
\verb$val m := Suc (Suc 0).$ \\
\verb$val odd := fun x. if x = Suc 0 then true else ?__.$ \\
\verb$val even := fun x. if x = Suc (Suc 0) || x = 0 then true else ?__.$\kern-200mm
\end{quote}
%
The partial model gives sufficient information to the user to evaluate the
goal: ``2 is even if 1 is odd, 1 is odd if 0 is even, and 0 is even.''
Our experience with the Nitpick~\cite{blanchette-nipkow-2010} model finder
is that users are mostly interested in the values assigned to uninterpreted
constants (e.g., \texttt{m}). Occasionally, the models of recursive functions
are instructive in the face of underspecification. A typical example is
the \texttt{head} function that returns the first element of a nonempty list:
%
\begin{quote}
\verb$data list A := Nil | Cons A (list A).$ \\[2\jot]
\verb$rec head : pi A. list A -> A :=$ \\
\verb$  forall y ys. head (Cons y ys) = y.$ \\[2\jot]
\verb$goal ~ (head Nil = 0).$
\end{quote}
%
Nunchaku transforms the specification of \verb$head$ into
%
\begin{quote}
\verb$head xs = match xs with Nil -> head xs | Cons y ys -> y end$
\end{quote}
%
where the unspecified \verb$Nil$ case is expressed via nonterminating
recursion (\verb$head$ \verb$xs$ \verb$=$ \verb$head$ \verb$xs$).
The tool exhibits a model in which \verb$head$ \verb$Nil$ is interpreted as
a nonzero value.

@@

\begin{description}
%{{{
  \item[Type inference] infers types and checks definitions;
%  \item[Type skolemization]
%    replaces $\exists \alpha.\; \varphi[\alpha]$
%      with $\varphi[\tau]$, where $\tau$ is a fresh type;
  \item[Monomorphization]
    specializes polymorphic definitions on their type arguments and removes
    unused definitions;
  \item[Elimination of equations]
    translates multiple-equation definitions of recursive functions into
      a single nested pattern matching;
  \item[Specialization]
    creates instances of functions with static arguments (i.e., an argument
    that is passed unchanged to all recursive calls);
  \item[Polarization]
    specializes predicates into a version used in positive positions
      and a version used in negative positions;
  \item[Unrolling]
    adds a decreasing argument to possibly ill-founded predicates; % to make them well founded;
  \item[Skolemization]
    introduces Skolem symbols for term variables;
  \item[Elimination of (co)inductive predicates]
    recasts a multiple-clause (co)inductive predicate specification
      into a recursive equation;
  \item[Elimination of higher-order constructs]
    eliminates $\lambda$-abstractions and substitutes arrays for
    higher-order functions;
  \item[Elimination of recursion]
    performs the encoding from Section~\ref{sec:encoding};
  \item[Elimination of pattern matching]
    rewrites pattern-matching expressions using datatype
    discriminators and selectors;
  \item[CVC4 invocation] runs CVC4 to obtain a model.
%}}}
\end{description}

>>>>

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying our translation, described in
Section~\ref{sec:encoding}. It is useful to connect it to the
standard notion of \emph{well-founded} function definitions, often called
\emph{terminating} definitions. % in a slight abuse of terminology.
In such definitions, all recursive calls are decreasing with respect to a well-founded
relation, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well founded, it is sufficient to require that
the defined function be terminating when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\,
\con{p}( x ) \teq \lite\bigl( x \leq 0,\allowbreak\;  1,\allowbreak\;
  2 * \con{p}( x - 1 ) \bigr)$,
where the theory~$T$ is integer arithmetic extended with %uninterpreted
$\con p:\typeint \rightarrow \typeint$, can be shown to be well founded
under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}, where theories such as
integer arithmetic can be axiomatized.

@@

Another useful class of function definitions is that of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be ill founded, without
their being inconsistent. 
Intuitively, productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
Given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the function defined by
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
%%% If what follows ':' is a complete sentence, the standard Am.E. convention
%%% is to capitalize it. --JB
falls within this class: Each call to $\const{e}$ produces one
constructor before entering the next call. Like terminating recursion,
productive corecursion totally specifies the functions it defines,
and the proof of Theorem~\ref{thm:adm} can be adapted to cover that case:

\begin{theorem}\label{thm:adm-co}
If $\DDD$ is a set of productive function definitions for\/
$\sfundefs{\Sigma}$, then it is admissible.
\end{theorem}

%Theorem~\ref{thm:adm} can be extended to
%cover such specifications, based on the observation that unfolding
%a corecursive definition infinitely computes a unique infinite object.
%We conjecture that a large fragment of mixed recursive--corecursive
%specifications are

@@

It is even possible to mix recursion and corecursion in the same total function
\cite{blanchette-et-al-2015-fouco}.
Beyond totality, an admissible set can contain underspecified functions
such as $\forallf{\con{f}} x : \typeint.\allowbreak\;\, \con{f}( x )
\teq \con{f}( x )$ or $\forallf{\con{g}} x.\allowbreak\;\, \con{g}( x
) \teq \con{g}( x + 1 )$. The latter is problematic operationally, because
in general the closure of a formula~$\psi$ that depends on some term
$\con{f}(\con{a})$ is an infinite set
$\{\psi\} \cup \{\con {g}(\con{a}+k) \teq \con {g}(\con{a}+k+1) \mid k \geq 0\}$.
A similar issue arise with corecursive definitions specifying infinite
\emph{acyclic} objects, such as the $\con{e}$ function introduced above.
Nonetheless, admissibility is still useful if the problem does not refer to
$\con{g}$ or $\con{e}$, because it tells us that we can safely ignore these
functions altogether.
%
We conjecture that it is safe to ignore all tail-recursive calls (i.e., calls that
occupy the right-hand side of the definition, potentially under some $\lite$
branch) when establishing well-foundedness or productivity, without affecting
admissibility.

An example of an inadmissible set is
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$,
where~$T$ is integer arithmetic extended to a set of uninterpreted symbols
$\{\con f, \con g :\typeint \rightarrow \typeint,\, \ldots \}$.
The set is inadmissible because the formula $\ltrue$ is
closed under function expansion with respect to this set
(since $\con{f}$ does not occur in $\ltrue$),
and yet there is no model of $T$
satisfying $\con{f}$'s definition. A more subtle example is
$\badassex$.
While this set has a model where $\con f$ and $\con g$ are interpreted as the
constant function $0$, it is not admissible since %the formula
$\con{f}( 0 ) \teq 1$ is
closed under function expansion \begin{rep}with respect to this set\end{rep}
and yet there exists
no interpretation satisfying both $\con{f}( 0 ) \teq 1$ and $\con{g}$'s
definition.

@@

The input syntax was inspired by that other systems based on higher-order
logic (e.g., Isabelle/HOL) and by functional programming languages (e.g.,
OCaml). The following simple problem gives a taste of the syntax:
%
\begin{quote}
\verb$data nat := 0 | Suc nat.$ \\[2\jot]
\verb$pred even : nat -> prop :=$ \\
\verb$  even 0;$ \\
\verb$  forall n. odd n => even (Suc n)$ \\
\verb$and odd : nat -> prop :=$ \\
\verb$  forall n. even n => odd (Suc n).$ \\[2\jot]
\verb$val m : nat.$ \\
\verb$goal even m && ~ (m = 0).$
\end{quote}
%
The problem defines a datatype (\texttt{nat}) and two mutually recursive
inductive predicates (\texttt{even} and \texttt{odd}), it declares a constant
\texttt{m}, and specifies a goal to satisfy (``\texttt{m} is even and nonzero'').
Nunchaku quickly finds the following partial model:
%
\begin{quote}
\verb$val m := Suc (Suc 0).$ \\
\verb$val odd := fun x. if x = Suc 0 then true else ?__.$ \\
\verb$val even := fun x. if x = Suc (Suc 0) || x = 0 then true else ?__.$\kern-200mm
\end{quote}
%
The partial model gives sufficient information to the user to evaluate the
goal: ``2 is even if 1 is odd, 1 is odd if 0 is even, and 0 is even.''
Our experience with the Nitpick~\cite{blanchette-nipkow-2010} model finder
is that users are mostly interested in the values assigned to uninterpreted
constants (e.g., \texttt{m}). Occasionally, the models of recursive functions
are instructive in the face of underspecification. A typical example is
the \texttt{head} function that returns the first element of a nonempty list:
%
\begin{quote}
\verb$data list A := Nil | Cons A (list A).$ \\[2\jot]
\verb$rec head : pi A. list A -> A :=$ \\
\verb$  forall y ys. head (Cons y ys) = y.$ \\[2\jot]
\verb$goal ~ (head Nil = 0).$
\end{quote}
%
Nunchaku transforms the definition of \verb$head$ into
%
\begin{quote}
\verb$head xs = match xs with Nil -> head xs | Cons y ys -> y end$
\end{quote}
%
where the unspecified \verb$Nil$ case is expressed via nonterminating
recursion (\verb$head$ \verb$xs$ \verb$=$ \verb$head$ \verb$xs$).
The tool exhibits a model in which \verb$head$ \verb$Nil$ is interpreted as
a nonzero value.

@@

\begin{description}
%{{{
  \item[Type inference] infers types and checks definitions;
%  \item[Type skolemization]
%    replaces $\exists \alpha.\; \varphi[\alpha]$
%      with $\varphi[\tau]$, where $\tau$ is a fresh type;
  \item[Monomorphization]
    specializes polymorphic definitions on their type arguments and removes
    unused definitions;
  \item[Elimination of equations]
    translates multiple-equation definitions of recursive functions into
      a single nested pattern matching;
  \item[Specialization]
    creates instances of functions with static arguments (i.e., an argument
    that is passed unchanged to all recursive calls);
  \item[Polarization]
    specializes predicates into a version used in positive positions
      and a version used in negative positions;
  \item[Unrolling]
    adds a decreasing argument to possibly ill-founded predicates; % to make them well founded;
  \item[Skolemization]
    introduces Skolem symbols for term variables;
  \item[Elimination of (co)inductive predicates]
    recasts a multiple-clause (co)inductive predicate definition
      into a recursive equation;
  \item[Elimination of higher-order constructs]
    eliminates $\lambda$-abstractions and substitutes arrays for
    higher-order functions;
  \item[Elimination of recursion]
    performs the encoding from Section~\ref{sec:encoding};
  \item[Elimination of pattern matching]
    rewrites pattern-matching expressions using datatype
    discriminators and selectors;
  \item[CVC4 invocation] runs CVC4 to obtain a model.
%}}}
\end{description}
